---
title: "Livestock Growth Modelling Analysis"
author: "Yusuf Omotosho"
date: "2024-05-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Project Overview ###

This project was conducted for the Department of Animal Breeding & Genetics at the University of Ibadan. The study focused on modeling livestock growth performance using non-linear regression models to analyze growth patterns of Holstein Friesian and F1 crossbred cattle. The analysis provided insights for improved livestock management.

**Objectives:**

- Assess growth performance using non-linear regression models.
-	Estimate model parameters and compare growth patterns.
-	Provide data-driven recommendations for livestock management.

**Tools & Libraries Used**

-	R Packages: nls, minpack.lm, ggplot2, dplyr, readxl
-	Software: R, Excel
-	Statistical Methods: AIC, BIC, R-squared

```{r, include=FALSE, echo=FALSE}
# Set CRAN mirror
options(repos = c(CRAN = "https://cloud.r-project.org/"))
```

### Install and Load Packages ###

```{r, include=FALSE}
# Step 1 : Install necessary packages
install.packages(c("minpack.lm", "tidyverse", "readxl"))
```

```{r}
# STEP 2 : Load necessary libraries
library(nlme)
library(minpack.lm)
library(ggplot2)
library(dplyr)
library(readxl)
library(tidyr)
```

### Data and Methodology ###

**1. Data Source and Structure :**

The dataset used for this analysis was imported from an Excel file (Org_Dataset.xlsx) containing observations on cattle growth. Key variables include:
-	'Age_Months': Age of the cattle in months.
-	'Weight_Kg': Weight of the cattle in kilograms.
-	'Breed_Group': Classification into Holstein Friesian or F1 crossbred.

```{r}
# Load dataset
cattle_data <- read_excel("cattle_dataset.xlsx")

# View the imported data
head(cattle_data) 

# Inspect data structure
str(cattle_data)

# Convert Breed_Group to a factor
cattle_data$Breed_Group <- as.factor(cattle_data$Breed_Group)

# Summarize breed types and their occurrences
breed_summary <- cattle_data %>%
  count(Breed_Group, name = "Occurrences")

# Display the summary
breed_summary
```

**2. Data Quality Checks**

```{r}
# Display any rows with non-finite values for diagnostic purposes
non_finite_rows <- cattle_data %>%
  filter(
    !is.finite(Age_Months) |  # Check for non-finite Age
    !is.finite(Weight_Kg) # Check for non-finite Weight
  )

if (nrow(non_finite_rows) > 0) {
  message("Non-finite values detected. Below are the rows with issues:")
  print(non_finite_rows)
} else {
  message("No non-finite values detected.")
}

# Check for missing values
cat("Missing Values:\n")
print(colSums(is.na(cattle_data)))

# Check for duplicate rows
cat("Duplicate Rows:\n")
print(nrow(cattle_data) - nrow(distinct(cattle_data)))

# Check for outliers using Z-score method
z_scores_Weight <- scale(cattle_data$Weight_Kg)
outliers_Weight <- which(abs(z_scores_Weight) > 3)
cat("Outliers Detected (Weight):\n")
print(outliers_Weight)

z_scores_age <- scale(cattle_data$Age_Months)
outliers_age <- which(abs(z_scores_age) > 3)
cat("Outliers Detected (Age):\n")
print(outliers_age)
```

**3. Statistical Summary of Dataset**

```{r}
# Summary statistics for each variable
cat("Summary Statistics:\n")
print(summary(cattle_data))
```

### Growth Curve Modeling ###

**Theoretical Basis and Equations**

1. 'Brody Model': Y(t)=A(1-B.e^(-kt))
2. 'Von Bertalanffy Model': Y(t)= A(1-B.e^(-kt))3
3. 'Logistic Model': Y(t) = A/(1+B.e^(-kt))

**Where:**

- 'Y(t)' is the population mean at time t.
- 'A' is the asymptotic weight or maximum value.
- 'B' is the proportion of asymptotic weight to be gained after birth.
- 'k' is the maturity rate.
- 't' is time.

**1. Model Fitting**

```{r}
# Define growth model equations
brody_model <- function(Age_Months, A, B, k) A * (1 - B * exp(-k * Age_Months))
vonB_model <- function(Age_Months, A, B, k) A * (1 - exp(-k * Age_Months))^B
log_model <- function(Age_Months, A, B, k) A / (1 + B * exp(-k * Age_Months))
```

**2. Fit models by Breed Type**

```{r}
# Fit models for each group
fit_model <- function(model_func, data) {
  nlsLM(Weight_Kg ~ model_func(Age_Months, A, B, k), 
        data = cattle_data,
        start = list(A = 730, B = 0.5, k = 0.01),
        control = nls.lm.control(maxiter = 1000))
}


fits <- cattle_data %>% 
  group_by(Breed_Group) %>% 
  summarise(
    brody = list(fit_model(brody_model, pick(everything()))),
    vonB = list(fit_model(vonB_model, pick(everything()))),
    log = list(fit_model(log_model, pick(everything())))
  )
print(fits)
```


**3. Parameter Estimation**

```{r}
extract_params <- function(model) {
  summary(model)$coefficients
}

params <- fits %>% 
  mutate(
    brody_params = lapply(brody, extract_params),
    vonB_params = lapply(vonB, extract_params),
    log_params = lapply(log, extract_params)
  )

params_upd <- as.data.frame(lapply(params, function(x) {
  gsub("[^[:print:]]", "", x)
}))

# Extract only the required parameter columns
params_selected <- params_upd %>%
  select(brody_params, vonB_params, log_params) %>%     # Keep only parameter columns
  mutate(
    brody_params = sapply(brody_params, paste, collapse = ", "),  # Convert lists to strings
    vonB_params  = sapply(vonB_params, paste, collapse = ", "),  
    log_params   = sapply(log_params, paste, collapse = ", ")   
  )

# Print the filtered and formatted results
print(params_selected)
```


### Model Comparison ###

**Assessment Criteria**
- 'AIC (Akaike Information Criterion)
- 'BIC' (Bayesian Information Criterion)
- 'R-squared'


**1. Retrieve residuals and fitted values:**

```{r}
compare_models <- function(model) {
  # Handle non-converged models
  if (!model$convInfo$isConv) {
    return(c(AIC = NA, BIC = NA, R2 = NA))
  }
  
  # Extract residuals and fitted values using functions
  residuals <- model$m$resid()  # Residuals
  fitted <- model$m$fitted()   # Fitted values
  
  # Calculate actual values as fitted + residuals
  actuals <- fitted + residuals

  # Compute metrics
  AIC_val <- AIC(model)
  BIC_val <- BIC(model)
  R2_val <- 1 - sum(residuals^2) / sum((actuals - mean(actuals))^2)

  return(c(AIC = AIC_val, BIC = BIC_val, R2 = R2_val))
}

# Extract metrics for each model and ensure numeric values are displayed
model_metrics <- fits %>% 
  rowwise() %>% 
  mutate(
    brody_metrics = list(as.numeric(compare_models(brody))),  # Convert to numeric vector
    vonB_metrics  = list(as.numeric(compare_models(vonB))),
    log_metrics   = list(as.numeric(compare_models(log)))
  ) %>%
  unnest_wider(brody_metrics, names_sep = "_") %>%  # Flatten metrics into separate columns
  unnest_wider(vonB_metrics, names_sep = "_") %>%
  unnest_wider(log_metrics, names_sep = "_") %>%
  ungroup()

# Print the cleaned-up metrics
print(model_metrics)
```


**2. Check for Non-convergence**

```{r}
# Function to check convergence
check_convergence <- function(model) {
  if (is.null(model)) return(data.frame(Status = "Error: Model is NULL"))
  
  conv_info <- tryCatch(model$convInfo, error = function(e) NULL)
  if (is.null(conv_info)) return(data.frame(Status = "Error: No convergence info"))
  
  residuals <- tryCatch(residuals(model), error = function(e) NA)
  fitted <- tryCatch(fitted(model), error = function(e) NA)
  actuals <- fitted + residuals
  
  # Calculate R2 if residuals and fitted values are valid
  R2 <- if (!anyNA(residuals) && !anyNA(fitted)) {
    1 - sum(residuals^2) / sum((actuals - mean(actuals))^2)
  } else {
    NA
  }
  
  # Return details as a data frame
  data.frame(
    Status = if (conv_info$isConv) "Converged" else "Not Converged",
    Iterations = conv_info$finIter,
    Tolerance = conv_info$finTol,
    StopCode = conv_info$stopCode,
    StopMessage = conv_info$stopMessage,
    R2 = R2
  )
}

# Extract coefficients
extract_coefficients <- function(model) {
  tryCatch(as.numeric(coef(model)), error = function(e) NA)
}

# Apply convergence checks and extract details
convergence_results <- fits %>%
  rowwise() %>%
  mutate(
    brody_check = list(check_convergence(brody)),
    vonB_check = list(check_convergence(vonB)),
    log_check = list(check_convergence(log)),
    brody_coeff = list(extract_coefficients(brody)),
    vonB_coeff = list(extract_coefficients(vonB)),
    log_coeff = list(extract_coefficients(log))
  ) %>%
  unnest_wider(brody_check, names_sep = "_") %>%  # Flatten convergence details
  unnest_wider(vonB_check, names_sep = "_") %>%
  unnest_wider(log_check, names_sep = "_") %>%
  unnest_longer(brody_coeff, values_to = "Brody_Param") %>% # Convert coefficients to rows
  unnest_longer(vonB_coeff, values_to = "vonB_Param") %>%
  unnest_longer(log_coeff, values_to = "Log_Param") %>%
  ungroup()

# Print results
print(convergence_results)
```


### Visualization ###

```{r}
plot_growth <- function(model_func, model, data, title) {
  # Prepare prediction data
  pred_data <- data.frame(Age_Months = seq(min(data$Age_Months), max(data$Age_Months), length.out = 100))
  pred_data$Weight_Kg <- predict(model, newdata = pred_data)
  
  # Create plot
  ggplot(data, aes(x = Age_Months, y = Weight_Kg, color = Breed_Group)) +
    geom_point(size = 2, alpha = 0.7) +
    geom_smooth(method = "loess", se = FALSE, color = "grey", linetype = "dashed") +
    geom_line(data = pred_data, aes(x = Age_Months, y = Weight_Kg), linewidth = 1.2, color = "blue") +
    labs(
      title = title, 
      subtitle = "Fitted Growth Model with Observed Data",
      x = "Age (months)", 
      y = "Weight (kg)"
    ) +
    theme_minimal(base_size = 14) +
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold"),
      plot.subtitle = element_text(hjust = 0.5),
      legend.position = "bottom"
    )
}

plots <- fits %>% 
  rowwise() %>% 
  mutate(
    brody_plot = list(plot_growth(brody_model, brody, cattle_data, paste((Group = factor(Breed_Group)), "- Brody Model"))),
    vonB_plot = list(plot_growth(vonB_model, vonB, cattle_data, paste((Group = factor(Breed_Group)), "- Von Bertalanffy Model"))),
    log_plot = list(plot_growth(log_model, log, cattle_data, paste((Group = factor(Breed_Group)), "- Logistic Model")))
  )
```


### Output plots ###

```{r}
for (plot_data in plots$brody_plot) print(plot_data)
for (plot_data in plots$vonB_plot) print(plot_data)
for (plot_data in plots$log_plot) print(plot_data)
```